---
title: "GAMM Model Comparison: Parallel vs Divergent Group Effects"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: show
    fig_width: 10
    fig_height: 6
---

<!-- To render this file from a PowerShell terminal:
  $env:RSTUDIO_PANDOC = "C:\Program Files\RStudio\resources\app\bin\quarto\bin\tools"
  & "C:\Program Files\R\R-4.5.2\bin\Rscript.exe" -e "rmarkdown::render('13_gamm_comparison_R.Rmd')"
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  dpi = 150,
  fig.path = "13_gamm_comparison_R_figures/"
)
set.seed(42)
```

# 1. Setup

```{r packages}
library(tidyverse)
library(mgcv)
library(gratia)

# Project-wide group colors (match notebooks 02/03 and 15_spaghetti_plots_R.Rmd)
group_cols <- c(
  "CONTROL" = "#377EB8",
  "GLAUCOMA - URAMOX" = "#E41A1C",
  "GLAUCOMA + URAMOX" = "#4DAF4A"
)
condition_cols <- c("CONTROL" = "#377EB8", "GLAUCOMA" = "#E41A1C")
contrast_cols <- c(
  "GLAUCOMA\u2212URAMOX vs CONTROL" = "#E41A1C",
  "GLAUCOMA+URAMOX vs CONTROL" = "#4DAF4A"
)

# Display labels for figures
LABEL_MAP <- c(
  "CONTROL"                        = "control",
  "GLAUCOMA - URAMOX"              = "glaucoma without oral acetazolamide",
  "GLAUCOMA + URAMOX"              = "glaucoma with oral acetazolamide",
  "GLAUCOMA"                       = "glaucoma",
  "GLAUCOMA\u2212URAMOX vs CONTROL" = "glaucoma without oral acetazolamide vs control",
  "GLAUCOMA+URAMOX vs CONTROL"     = "glaucoma with oral acetazolamide vs control"
)
```

```{r load-data}
df <- read_csv("results/eye_liquid_spread_measurements.csv") %>%
  mutate(
    group = factor(group, levels = c("CONTROL", "GLAUCOMA - URAMOX", "GLAUCOMA + URAMOX")),
    case  = factor(case),
    condition = factor(
      ifelse(grepl("GLAUCOMA", group), "GLAUCOMA", "CONTROL"),
      levels = c("CONTROL", "GLAUCOMA")
    )
  )
```

```{r data-summary}
df %>%
  group_by(group) %>%
  summarise(
    n_cases   = n_distinct(case),
    n_obs     = n(),
    time_min  = min(time_sec),
    time_max  = max(time_sec),
    area_mean = mean(area_cm2),
    area_sd   = sd(area_cm2),
    .groups   = "drop"
  ) %>%
  knitr::kable(digits = 3, caption = "Summary statistics by group")
```

# 2. Two-Group Comparison (CONTROL vs GLAUCOMA)

We fit two nested models that differ in how group affects the time trajectory:

- **Parallel**: `area ~ condition + s(time) + s(time, case, bs="fs")` — one master
  non-linear shape; groups are shifted vertically.
- **Divergent**: `area ~ condition + s(time, by=condition) + s(time, case, bs="fs")` —
  each group gets its own non-linear shape; no redundant global smooth.

Both models use ML for valid likelihood-based comparison.

## Fit Models

```{r fit-2g}
m_par_2g <- gam(
  area_cm2 ~ condition +
    s(time_sec, k = 20, bs = "cr") +
    s(time_sec, case, bs = "fs", k = 10),
  data = df, method = "ML"
)

m_div_2g <- gam(
  area_cm2 ~ condition +
    s(time_sec, by = condition, k = 20, bs = "cr") +
    s(time_sec, case, bs = "fs", k = 10),
  data = df, method = "ML"
)
```

## Model Summaries

```{r summary-par-2g}
summary(m_par_2g)
```

```{r summary-div-2g}
summary(m_div_2g)
```

## Parametric Coefficients (Intercepts)

The parametric terms capture the mean level for each group. In the Parallel model
the group coefficient is the *entire* constant vertical shift between curves;
in the Divergent model the group-specific smooths absorb shape differences, so
the coefficient reflects the *residual* mean-level offset after accounting for
different shapes.

```{r coef-2g}
ptab_par_2g <- as.data.frame(summary(m_par_2g)$p.table) %>%
  rownames_to_column("Term") %>%
  mutate(Model = "Parallel")
ptab_div_2g <- as.data.frame(summary(m_div_2g)$p.table) %>%
  rownames_to_column("Term") %>%
  mutate(Model = "Divergent")

coef_2g <- bind_rows(ptab_par_2g, ptab_div_2g) %>%
  select(Model, Term, Estimate, `Std. Error`, `t value`, `Pr(>|t|)`)

knitr::kable(coef_2g, digits = 4,
             caption = "Parametric coefficients — 2-group models")
```

```{r coef-interpret-2g, results='asis'}
get_coef <- function(model, term) {
  tab <- summary(model)$p.table
  list(est = tab[term, "Estimate"],
       se  = tab[term, "Std. Error"],
       p   = tab[term, "Pr(>|t|)"])
}

par_grp_2g <- get_coef(m_par_2g, "conditionGLAUCOMA")
div_grp_2g <- get_coef(m_div_2g, "conditionGLAUCOMA")

cat("### Intercept Conclusions — 2-Group\n\n")
cat(sprintf("- **Parallel model**: GLAUCOMA shifts area by **%.3f cm\u00b2** ",
            par_grp_2g$est))
cat(sprintf("(SE = %.3f, p = %.4f) relative to CONTROL.\n", par_grp_2g$se, par_grp_2g$p))
cat(sprintf("- **Divergent model**: GLAUCOMA shifts area by **%.3f cm\u00b2** ",
            div_grp_2g$est))
cat(sprintf("(SE = %.3f, p = %.4f) relative to CONTROL.\n\n", div_grp_2g$se, div_grp_2g$p))

shift_change <- abs(par_grp_2g$est - div_grp_2g$est)
cat(sprintf("The group intercept changes by %.3f cm\u00b2 between models. ", shift_change))
if (shift_change < 0.1 * abs(par_grp_2g$est)) {
  cat("This is a small change (&lt; 10% of the Parallel estimate), indicating that ",
      "the baseline shift is robust regardless of whether group-specific shapes are ",
      "allowed.\n\n")
} else {
  cat("This is a notable change, suggesting that allowing different shapes absorbs ",
      "some of what the Parallel model attributed to the intercept shift.\n\n")
}

if (par_grp_2g$p < 0.05 && div_grp_2g$p < 0.05) {
  cat("The group intercept is **significant in both models** (p &lt; 0.05), confirming ",
      "that GLAUCOMA and CONTROL differ in their mean area level.\n")
} else if (par_grp_2g$p < 0.05 && div_grp_2g$p >= 0.05) {
  cat("The group intercept is **significant in the Parallel model but not in the ",
      "Divergent model**. This suggests that what appears as a level shift in the ",
      "Parallel model is partly explained by different curve shapes in the Divergent model.\n")
} else if (par_grp_2g$p >= 0.05 && div_grp_2g$p >= 0.05) {
  cat("The group intercept is **not significant in either model** (p &ge; 0.05). ",
      "There is no evidence of a mean-level difference between groups.\n")
} else {
  cat("The group intercept is non-significant in the Parallel model but significant ",
      "in the Divergent model &mdash; an unusual pattern that warrants careful inspection.\n")
}
```

## AIC Comparison

$$\text{AIC} = 2k - 2\ln(L)$$

where $L$ is the maximum likelihood and $k$ is the effective degrees of freedom.
Lower AIC is better; $\Delta\text{AIC} > 2$ indicates a meaningful difference,
$> 10$ indicates strong evidence.

```{r aic-2g}
aic_2g <- AIC(m_par_2g, m_div_2g) %>%
  as.data.frame() %>%
  rownames_to_column("model") %>%
  mutate(
    model = c("Parallel", "Divergent"),
    delta_AIC = AIC - min(AIC)
  )

knitr::kable(aic_2g, digits = 1,
             caption = "AIC comparison — 2-group (CONTROL vs GLAUCOMA)")
```

## Likelihood Ratio Test

$$D = -2 \ln\!\left(\frac{L_{\text{parallel}}}{L_{\text{divergent}}}\right) \sim \chi^2$$

The Parallel model is nested inside the Divergent model (it is the Divergent model
with the constraint that both group shapes are identical). A significant p-value
means the Divergent model fits significantly better.

```{r lrt-chisq-2g}
lrt_chisq_2g <- anova(m_par_2g, m_div_2g, test = "Chisq")
print(lrt_chisq_2g)
```

The F-test is more conservative than Chi-squared and is recommended for
Gaussian models where the residual variance is estimated rather than known.

```{r lrt-f-2g}
lrt_f_2g <- anova(m_par_2g, m_div_2g, test = "F")
print(lrt_f_2g)
```

## Interpretation

```{r interpret-2g, results='asis'}
aic_diff_2g <- AIC(m_par_2g) - AIC(m_div_2g)
lrt_p_chisq_2g <- lrt_chisq_2g[["Pr(>Chi)"]][2]
lrt_p_f_2g     <- lrt_f_2g[["Pr(>F)"]][2]

cat("### Two-Group Result\n\n")
cat(sprintf("- **&Delta;AIC** (Parallel &minus; Divergent) = %.1f\n", aic_diff_2g))
cat(sprintf("- **LRT p-value (Chi-sq)** = %.4f\n", lrt_p_chisq_2g))
cat(sprintf("- **LRT p-value (F-test)** = %.4f\n\n", lrt_p_f_2g))

chisq_sig <- !is.na(lrt_p_chisq_2g) && lrt_p_chisq_2g < 0.05
f_sig     <- !is.na(lrt_p_f_2g) && lrt_p_f_2g < 0.05
aic_sig   <- aic_diff_2g > 2

if (chisq_sig && f_sig && aic_sig) {
  cat("**All three criteria favour the Divergent model.** The non-linear spreading ",
      "dynamics differ significantly between CONTROL and GLAUCOMA groups ",
      "&mdash; the difference is not merely a baseline shift.\n")
} else if (!chisq_sig && !f_sig && !aic_sig) {
  cat("**No criterion favours the Divergent model.** The Parallel model ",
      "is adequate: groups differ by a vertical shift, but their non-linear ",
      "spreading dynamics follow the same shape.\n")
} else {
  cat("**Mixed results.** The criteria do not fully agree:\n\n")
  cat(sprintf("- AIC: %s (|&Delta;AIC| = %.1f)\n",
              ifelse(aic_sig, "favours Divergent", "no meaningful difference"), abs(aic_diff_2g)))
  cat(sprintf("- Chi-squared LRT: %s (p = %.4f)\n",
              ifelse(chisq_sig, "significant", "not significant"), lrt_p_chisq_2g))
  cat(sprintf("- F-test LRT: %s (p = %.4f)\n\n",
              ifelse(f_sig, "significant", "not significant"), lrt_p_f_2g))
  cat("When tests disagree, the more conservative F-test should be given more weight.\n")
}
```

## Predicted Curves

```{r pred-2g, fig.height=5}
time_grid <- seq(min(df$time_sec), max(df$time_sec), length.out = 200)

pred_2g <- expand_grid(
  time_sec  = time_grid,
  condition = levels(df$condition)
) %>%
  mutate(
    condition = factor(condition, levels = levels(df$condition)),
    case = levels(df$case)[1]
  )

pred_2g$fit_par <- predict(m_par_2g, newdata = pred_2g, type = "response",
                           exclude = "s(time_sec,case)")
pred_2g$fit_div <- predict(m_div_2g, newdata = pred_2g, type = "response",
                           exclude = "s(time_sec,case)")

pred_2g_long <- pred_2g %>%
  select(time_sec, condition, fit_par, fit_div) %>%
  pivot_longer(
    cols = c(fit_par, fit_div),
    names_to = "model",
    values_to = "fit"
  ) %>%
  mutate(model = recode(model,
    "fit_par" = "Parallel",
    "fit_div" = "Divergent"
  ))

ggplot(pred_2g_long, aes(x = time_sec, y = fit, colour = condition)) +
  geom_line(linewidth = 1) +
  facet_wrap(~model) +
  scale_colour_manual(values = condition_cols, labels = LABEL_MAP) +
  labs(
    x = "Time (s)",
    y = expression(Area ~ (cm^2)),
    colour = "Group"
  ) +
  theme_minimal(base_size = 13)
```

## Group Difference — Divergent Model

```{r diff-2g, fig.height=5}
Xp_2g <- predict(m_div_2g, newdata = pred_2g, type = "lpmatrix",
                 exclude = "s(time_sec,case)")
Vcov_2g <- vcov(m_div_2g)

idx_ctrl_2g <- which(pred_2g$condition == "CONTROL")
idx_glau_2g <- which(pred_2g$condition == "GLAUCOMA")

Xp_diff_2g <- Xp_2g[idx_glau_2g, ] - Xp_2g[idx_ctrl_2g, ]

diff_2g <- tibble(
  time_sec   = time_grid,
  difference = pred_2g$fit_div[idx_glau_2g] - pred_2g$fit_div[idx_ctrl_2g],
  se         = sqrt(rowSums((Xp_diff_2g %*% Vcov_2g) * Xp_diff_2g))
) %>%
  mutate(
    lower = difference - 1.96 * se,
    upper = difference + 1.96 * se
  )

ggplot(diff_2g, aes(x = time_sec, y = difference)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, fill = condition_cols["GLAUCOMA"]) +
  geom_line(linewidth = 1, colour = condition_cols["GLAUCOMA"]) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  labs(
    x = "Time (s)",
    y = expression(Delta ~ Area ~ (cm^2)),
    title = "glaucoma \u2212 control difference (Divergent model)",
    subtitle = "95% CI via linear predictor matrix (Wood 2017, \u00a77.2.5)"
  ) +
  theme_minimal(base_size = 13)
```


# 3. Three-Group Comparison

Same approach with the full three-level group factor:
CONTROL vs GLAUCOMA − URAMOX vs GLAUCOMA + URAMOX.

## Fit Models

```{r fit-3g}
m_par_3g <- gam(
  area_cm2 ~ group +
    s(time_sec, k = 20, bs = "cr") +
    s(time_sec, case, bs = "fs", k = 10),
  data = df, method = "ML"
)

m_div_3g <- gam(
  area_cm2 ~ group +
    s(time_sec, by = group, k = 20, bs = "cr") +
    s(time_sec, case, bs = "fs", k = 10),
  data = df, method = "ML"
)
```

## Model Summaries

```{r summary-par-3g}
summary(m_par_3g)
```

```{r summary-div-3g}
summary(m_div_3g)
```

## Parametric Coefficients (Intercepts)

```{r coef-3g}
ptab_par_3g <- as.data.frame(summary(m_par_3g)$p.table) %>%
  rownames_to_column("Term") %>%
  mutate(Model = "Parallel")
ptab_div_3g <- as.data.frame(summary(m_div_3g)$p.table) %>%
  rownames_to_column("Term") %>%
  mutate(Model = "Divergent")

coef_3g <- bind_rows(ptab_par_3g, ptab_div_3g) %>%
  select(Model, Term, Estimate, `Std. Error`, `t value`, `Pr(>|t|)`)

knitr::kable(coef_3g, digits = 4,
             caption = "Parametric coefficients — 3-group models")
```

```{r coef-interpret-3g, results='asis'}
grp_terms <- c("groupGLAUCOMA - URAMOX", "groupGLAUCOMA + URAMOX")
grp_labels <- c("GLAUCOMA\u2212URAMOX", "GLAUCOMA+URAMOX")

cat("### Intercept Conclusions — 3-Group\n\n")

for (i in seq_along(grp_terms)) {
  par_c <- get_coef(m_par_3g, grp_terms[i])
  div_c <- get_coef(m_div_3g, grp_terms[i])

  cat(sprintf("**%s vs CONTROL:**\n\n", grp_labels[i]))
  cat(sprintf("- Parallel: **%.3f cm\u00b2** (SE = %.3f, p = %.4f)\n",
              par_c$est, par_c$se, par_c$p))
  cat(sprintf("- Divergent: **%.3f cm\u00b2** (SE = %.3f, p = %.4f)\n",
              div_c$est, div_c$se, div_c$p))

  shift_change <- abs(par_c$est - div_c$est)
  if (shift_change < 0.1 * max(abs(par_c$est), 1e-6)) {
    cat(sprintf("- Intercept shift is stable across models (\u0394 = %.3f cm\u00b2).\n",
                shift_change))
  } else {
    cat(sprintf("- Intercept changes notably between models (\u0394 = %.3f cm\u00b2), ",
                shift_change))
    cat("suggesting shape differences absorb part of the level shift.\n")
  }

  if (par_c$p < 0.05 && div_c$p < 0.05) {
    cat("- Significant in **both** models: robust mean-level difference from CONTROL.\n\n")
  } else if (par_c$p < 0.05 && div_c$p >= 0.05) {
    cat("- Significant only in the **Parallel** model: the apparent level shift is ",
        "partly explained by different curve shapes.\n\n")
  } else if (par_c$p >= 0.05 && div_c$p >= 0.05) {
    cat("- **Not significant** in either model: no evidence of a mean-level difference.\n\n")
  } else {
    cat("- Significant only in the **Divergent** model: warrants careful inspection.\n\n")
  }
}
```

## AIC Comparison

```{r aic-3g}
aic_3g <- AIC(m_par_3g, m_div_3g) %>%
  as.data.frame() %>%
  rownames_to_column("model") %>%
  mutate(
    model = c("Parallel", "Divergent"),
    delta_AIC = AIC - min(AIC)
  )

knitr::kable(aic_3g, digits = 1,
             caption = "AIC comparison — 3-group")
```

## Likelihood Ratio Test

```{r lrt-chisq-3g}
lrt_chisq_3g <- anova(m_par_3g, m_div_3g, test = "Chisq")
print(lrt_chisq_3g)
```

```{r lrt-f-3g}
lrt_f_3g <- anova(m_par_3g, m_div_3g, test = "F")
print(lrt_f_3g)
```

## Interpretation

```{r interpret-3g, results='asis'}
aic_diff_3g <- AIC(m_par_3g) - AIC(m_div_3g)
lrt_p_chisq_3g <- lrt_chisq_3g[["Pr(>Chi)"]][2]
lrt_p_f_3g     <- lrt_f_3g[["Pr(>F)"]][2]

cat("### Three-Group Result\n\n")
cat(sprintf("- **&Delta;AIC** (Parallel &minus; Divergent) = %.1f\n", aic_diff_3g))
cat(sprintf("- **LRT p-value (Chi-sq)** = %.4f\n", lrt_p_chisq_3g))
cat(sprintf("- **LRT p-value (F-test)** = %.4f\n\n", lrt_p_f_3g))

chisq_sig <- !is.na(lrt_p_chisq_3g) && lrt_p_chisq_3g < 0.05
f_sig     <- !is.na(lrt_p_f_3g) && lrt_p_f_3g < 0.05
aic_sig   <- aic_diff_3g > 2

if (chisq_sig && f_sig && aic_sig) {
  cat("**All three criteria favour the Divergent model.** At least one group has ",
      "non-linear spreading dynamics that differ from the others.\n")
} else if (!chisq_sig && !f_sig && !aic_sig) {
  cat("**No criterion favours the Divergent model.** The Parallel model ",
      "is adequate: all three groups follow the same non-linear shape, ",
      "differing only in baseline area.\n")
} else {
  cat("**Mixed results.** The criteria do not fully agree:\n\n")
  cat(sprintf("- AIC: %s (|&Delta;AIC| = %.1f)\n",
              ifelse(aic_sig, "favours Divergent", "no meaningful difference"), abs(aic_diff_3g)))
  cat(sprintf("- Chi-squared LRT: %s (p = %.4f)\n",
              ifelse(chisq_sig, "significant", "not significant"), lrt_p_chisq_3g))
  cat(sprintf("- F-test LRT: %s (p = %.4f)\n\n",
              ifelse(f_sig, "significant", "not significant"), lrt_p_f_3g))
  cat("When tests disagree, the more conservative F-test should be given more weight.\n")
}
```

## Predicted Curves

```{r pred-3g, fig.height=5}
pred_3g <- expand_grid(
  time_sec = time_grid,
  group    = levels(df$group)
) %>%
  mutate(
    group = factor(group, levels = levels(df$group)),
    case  = levels(df$case)[1]
  )

pred_3g$fit_par <- predict(m_par_3g, newdata = pred_3g, type = "response",
                           exclude = "s(time_sec,case)")
pred_3g$fit_div <- predict(m_div_3g, newdata = pred_3g, type = "response",
                           exclude = "s(time_sec,case)")

pred_3g_long <- pred_3g %>%
  select(time_sec, group, fit_par, fit_div) %>%
  pivot_longer(
    cols = c(fit_par, fit_div),
    names_to = "model",
    values_to = "fit"
  ) %>%
  mutate(model = recode(model,
    "fit_par" = "Parallel",
    "fit_div" = "Divergent"
  ))

ggplot(pred_3g_long, aes(x = time_sec, y = fit, colour = group)) +
  geom_line(linewidth = 1) +
  facet_wrap(~model) +
  scale_colour_manual(values = group_cols, labels = LABEL_MAP) +
  labs(
    x = "Time (s)",
    y = expression(Area ~ (cm^2)),
    colour = "Group"
  ) +
  theme_minimal(base_size = 13)
```

## Group Differences — Divergent Model

```{r diff-3g, fig.height=5}
Xp_3g <- predict(m_div_3g, newdata = pred_3g, type = "lpmatrix",
                 exclude = "s(time_sec,case)")
Vcov_3g <- vcov(m_div_3g)

idx_ctrl_3g  <- which(pred_3g$group == "CONTROL")
idx_minus_3g <- which(pred_3g$group == "GLAUCOMA - URAMOX")
idx_plus_3g  <- which(pred_3g$group == "GLAUCOMA + URAMOX")

Xp_diff_minus <- Xp_3g[idx_minus_3g, ] - Xp_3g[idx_ctrl_3g, ]
Xp_diff_plus  <- Xp_3g[idx_plus_3g, ]  - Xp_3g[idx_ctrl_3g, ]

diff_3g <- bind_rows(
  tibble(
    time_sec   = time_grid,
    difference = pred_3g$fit_div[idx_minus_3g] - pred_3g$fit_div[idx_ctrl_3g],
    se         = sqrt(rowSums((Xp_diff_minus %*% Vcov_3g) * Xp_diff_minus)),
    contrast   = "GLAUCOMA\u2212URAMOX vs CONTROL"
  ),
  tibble(
    time_sec   = time_grid,
    difference = pred_3g$fit_div[idx_plus_3g] - pred_3g$fit_div[idx_ctrl_3g],
    se         = sqrt(rowSums((Xp_diff_plus %*% Vcov_3g) * Xp_diff_plus)),
    contrast   = "GLAUCOMA+URAMOX vs CONTROL"
  )
) %>%
  mutate(
    lower = difference - 1.96 * se,
    upper = difference + 1.96 * se
  )

ggplot(diff_3g, aes(x = time_sec, y = difference, colour = contrast, fill = contrast)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  scale_colour_manual(values = contrast_cols, labels = LABEL_MAP) +
  scale_fill_manual(values = contrast_cols, labels = LABEL_MAP) +
  labs(
    x = "Time (s)",
    y = expression(Delta ~ Area ~ (cm^2)),
    colour = "Contrast",
    fill = "Contrast",
    title = "Group differences — Divergent model (3-group)",
    subtitle = "95% CI via linear predictor matrix (Wood 2017, \u00a77.2.5)"
  ) +
  theme_minimal(base_size = 13)
```


# 4. Diagnostics

```{r diag-par, fig.height=7}
cat("=== Parallel model (3-group) ===\n")
par(mfrow = c(2, 2))
gam.check(m_par_3g)
```

```{r diag-div, fig.height=7}
cat("=== Divergent model (3-group) ===\n")
par(mfrow = c(2, 2))
gam.check(m_div_3g)
```


# Session Info

```{r session-info}
sessionInfo()
```
